---
title: "Week 3 Update: Passive Ownership Measurement and Preliminary Volatility Analysis"
author: "Ryan Levy, Srimaan Bekkari, Ryan Rammacher, West Plowman"
date: "2026-02-12"
output:
  pdf_document: default
  html_document: default
---

# 1. Data Sources and Passive Ownership Measure

This section describes the institutional ownership data used to construct a proxy for passive investment. We use 13F filings obtained through the SEC API, focusing on holdings reported by BlackRock, Vanguard, and State Street.

To measure passive ownership at the firm–year level, we compute the proportion of publicly available shares held by these institutions.

$$
\text{PassiveOwnership}_{it} = 
\frac{\text{SharesHeldByPassive}_{it}}{\text{SharesOutstanding}_{it}}
$$

```{r}
# load in passive investment data
# gives shares held per reporting period per year aggrigated between BlackRock, Vanguard, and State Street
p_own_raw <- read.csv("13f_api_outputs/BVS_13f_holdings_panel.csv")

p_own <- p_own_raw
# format date column
p_own$report_date <- as.Date(p_own$report_date)
p_own$year <- as.integer(format(p_own$report_date, "%Y"))
# divides shares held by shares outstanding for proxy passive ownership metric
p_own$percentage_owned <- p_own$shares_held_total / p_own$shares_outstanding
```

# 2. Illustrative Example: Apple Inc. (AAPL)

To illustrate the construction of the passive ownership metric, we present a demonstration using Apple Inc. (AAPL). We aggregate institutional holdings by year and compute annual average passive ownership.

```{r}
# subset AAPL for demo
p_own_appl <- p_own[p_own$ticker == "AAPL",]

# consolidates averages per year
p_own_aapl_yr <- aggregate(percentage_owned ~ year, data = p_own_appl, FUN = mean)


head(p_own_aapl_yr)
```

# 3. Stock Price Data and Volatility Construction

We next introduce daily stock price data to construct a volatility measure that can be compared to passive ownership levels. Daily log returns are computed and aggregated into an annualized realized volatility measure.

```{r}
# load in data
aapl_raw <- read.csv("ipo_to_2026_stock_volatility/aapl_us_d.csv")
aapl <- aapl_raw
# format date
aapl$Data <- as.Date(aapl$Data)
aapl$year <- as.integer(format(aapl$Data, "%Y"))

# daily log return r_t = ln(Close_t / Close_{t-1})
aapl$log_ret <- c(NA, diff(log(aapl$Close)))

# year and annualized realized volatility
years <- sort(unique(aapl$year))
aapl_vol_annual <- data.frame(
  year = years,
  ann_vol = NA_real_,
  n_days = NA_integer_,
  stringsAsFactors = FALSE
)

# compute sigma_y = sd(daily returns in year y) * sqrt(252)
for (i in seq_along(years)) {
  y <- years[i]
  
  r <- aapl$log_ret[aapl$year == y]
  r <- r[!is.na(r)]
  
  aapl_vol_annual$ann_vol[i] <- sd(r) * sqrt(252)
  aapl_vol_annual$n_days[i]  <- length(r)
}

# Inspect
head(aapl_vol_annual, 10)
```
# 4. Merging Passive Ownership and Volatility Data

The passive ownership and volatility datasets are merged at the yearly level. This allows for direct comparison between changes in institutional ownership and realized stock volatility.

```{r}
merged_df <- merge(
  p_own_aapl_yr,
  aapl_vol_annual,
  by = "year",
  all = FALSE
)

merged_df[,c("year", "percentage_owned", "ann_vol")]

par(mar = c(5, 4, 4, 4) + 0.1)

plot(
  merged_df$year,
  merged_df$percentage_owned,
  type = "o",
  pch = 16,
  xlab = "Year",
  ylab = "Passive Ownership"
)

par(new = TRUE)

plot(
  merged_df$year,
  merged_df$ann_vol,
  type = "o",
  pch = 17,
  axes = FALSE,
  xlab = "",
  ylab = ""
)

axis(side = 4)
mtext("Annualized Volatility", side = 4, line = 3)
title("AAPL: Passive Ownership vs Annualized Volatility", line = 2)

legend(
  "topleft",
  legend = c("Passive Ownership", "Annualized Volatility"),
  pch = c(16, 17),
  lty = 1,
  bty = "n"
)



```
Aside from the large spike in passive ownership in 2014, the two lines seem to follow a similar pattern.

# 5. High-Frequency Visualization (Non-Annualized Data)

To examine the relationship in finer detail, we plot quarterly passive ownership alongside daily stock price data beginning in 2010.

```{r}
aapl_raw_2 <- aapl_raw
p_own_raw_2 <- p_own_raw

aapl_raw_2$Data <- as.Date(aapl_raw_2$Data)
p_own_raw_2$report_date <- as.Date(p_own_raw_2$report_date)

p_own_raw_2$percentage_owned <- p_own_raw_2$shares_held_total / p_own_raw_2$shares_outstanding

p_own_raw_2 <- p_own_raw_2[p_own_raw_2$ticker == "AAPL",]

head(aapl_raw_2)
head(p_own_raw_2)
```

```{r}
cut_date <- as.Date("2010-01-01")

aapl_2010 <- subset(aapl_raw_2, Data >= cut_date)
p_2010 <- subset(p_own_raw_2, ticker == "AAPL" & report_date >= cut_date)

plot(aapl_2010$Data, aapl_2010$Close, type="l",
     xlab="Date", ylab="Close Price",
     main="AAPL: Close vs Passive Ownership (2010+)")

par(new=TRUE)

plot(p_2010$report_date, p_2010$percentage_owned,
     type="l", lty=3, axes=FALSE, xlab="", ylab="")

axis(side=4)
mtext("Passive Ownership", side=4, line=3)

legend("topleft",
       legend = c("AAPL Close", "AAPL Passive Ownership"),
       lty = c(1, 3),
       bty = "n")
```

Two periods of sharp increases in passive ownership are visible and warrant further investigation. Perhaps also removing the linear trend from the data to only see the volatility would help.

```{r}
# compute log returns (centered around 0)
aapl_2010$log_ret <- c(NA, diff(log(aapl_2010$Close)))

plot(aapl_2010$Data, aapl_2010$log_ret, type="l",
     xlab="Date", ylab="Daily Log Return",
     main="AAPL Returns vs Passive Ownership (2010+)")

abline(h = 0, lty = 2)  # zero reference line

par(new=TRUE)

plot(p_2010$report_date, p_2010$percentage_owned,
     type="l", lty=3, axes=FALSE, xlab="", ylab="")

axis(side=4)
mtext("Passive Ownership", side=4, line=3)

legend("topleft",
       legend = c("AAPL Daily Returns", "Passive Ownership"),
       lty = c(1, 3),
       bty = "n")
```

It seems like large spikes of volatility follow the sharp increase in passive ownership, but further analysis is still to be done.